---
title: "Get started with dplyr"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



If you obtained a different (correct) answer than those listed on the solutions page, please feel free to post your answer as a comment on that page.

## Exercise 1
Install and load the package dplyr package. Given the metadata:

```{r}
library(dplyr)
df=data.frame(Theoph)
help(Theoph)

```



## Exercise 2

Use the `names()` function to get the column names of df.

## Exercise 3

Let's practice using the `select()` function. This allows you to work with just column names instead of indices.
a) Select only the columns starting from Subject to Dose
b) Only select the Wt and Dose columns now.

Learn more about dplyr in Section 5 Using `dplyr` on one and multiple Datasets of the online course R Data Pre-Processing & Data Management - Shape your Data! Rated 4.6 / 5 (45 ratings) . 473 students enrolled

## Exercise 4
Let's look at the sample with Dose greater than 5 mg/kg. Use the filter command() to return df with Dose>5'

## Exercise 5

Great. Now use filter command to return df with `Dose>5` and Time greater than the mean Time.

## Exercise 6

Now let's try sorting the data. Use the arrange() function to
1) arrange df by weight (descending)
2) arrange df by weight (ascending)
3) arrange df by weight (ascending) and Time (descending)

## Exercise 7

The `mutate()` command allows you to create a new column using conditions and data derived from other columns. Use `mutate()` command to create a new column called `trend` that equals to `Time-mean(Time)`. This will tell you how far each time value is from its mean. Set `na.rm=TRUE`.

## Exercise 8

Given the meta-data

76.2 kg Super-middleweight
72.57 kg Middleweight
69.85 kg Light-middleweight
66.68 kg Welterweight

Use the mutate function to classify the weight using the information above. For the purpose of this exercise, considering anything above 76.2 kg to be Super-middleweight and anything below 66.8 to be Welterweight. Anything below 76.2 to be middleweight and anything below 72.57 to be light-middleweight. Store the classifications under weight_cat. Hint: Use ifelse function() with mutate() to achieve this. Store this back into df.

## Exercise 9


 
Use the `groupby()` command to group df by weight_cat. This allows us to use aggregated functions similar to group by in SQL. Store this in a df called `weight_group`

## Exercise 10

Use the `summarize()` command on the `weight_group` created in Question 9 to find the mean Time and sum of `Dose` received by each weight 
categories.

