---
title: "Databases"
author: "Hicham Zmarrou"
date: "April 01, 2017"
output: html_document
---



### Exercises


1. Load libraries: dplyr, dbplyr, DBI, RSQLite, nycflights13. Create a connection to temporal in-memory SQLite database (the database will be created on-the-fly so you do not have to take care of this).

####################
#                  #
#    Exercise 1    #
#                  #
####################


```{r, message = TRUE}
library(dplyr)
library(dbplyr)
library(DBI)
library(RSQLite)
library(nycflights13)

con <- dbConnect(SQLite(), path = ":memory:")

```


####################
#                  #
#    Exercise 2    #
#                  #
####################

2. Upload data sets nycflights13::flights and nycflights13::planes to the database as non temporal tables

```{r, message = TRUE}
copy_to(con, flights, name = 'flights', temporary = FALSE)
copy_to(con, planes, name = 'planes', temporary = FALSE)

```


####################
#                  #
#    Exercise 3    #
#                  #
####################
3. List names of columns in flights table (hint: There is a function in DBI package for this).

```{r}
dbListFields(con, 'flights')

```

####################
#                  #
#    Exercise 4    #
#                  #
####################

4. Use SQL query to count number of flights per carrier and pull it to a local tibble.

```{r, message = TRUE} 
tbl(con, sql("select carrier, count(*) as count from flights group by 1")) %>%
  collect()

```

5. Do the same thing using dplyr verbs instead of SQL.
####################
#                  #
#    Exercise 5    #
#                  #
####################
```{r}
tbl(con, 'flights') %>%
  group_by(carrier) %>%
  summarise(count = n()) %>%
  collect()
```


6. Calculate number of flights, mean and total distance per plane, discard records with NA at tailnum column and save it to temporal table.

####################
#                  #
#    Exercise 6    #
#                  #
####################

```{r}
tbl(con, 'flights') %>%
  group_by(tailnum) %>%
  summarise(count=n(),
            mean_distance = mean(distance),
            total_distance = sum(distance)) %>%
  filter(!is.na(tailnum)) %>%
  compute(name = 'planes_distance')
  
```  


7. List all the tables in the database (hint: There is a function in DBI package for this).

####################
#                  #
#    Exercise 7    #
#                  #
####################
```{r}
dbListTables(con)
```



8. Use head(), tail() and nrow() to investigate table planes. Do you understand why the latter two do not work?

####################
#                  #
#    Exercise 8    #
#                  #
####################

```{r}
tbl(con, 'planes') %>%
  head()
  
  tbl(con, 'planes') %>%
  tail()
## Error: tail() is not supported by sql sources
tbl(con, 'planes') %>%
  nrow()
```

9.Join the table from exercise 6 with planes table and without pulling the data to local frame find manufacturers and models of 10 planes with the highest total flown distance.

####################
#                  #
#    Exercise 9    #
#                  #
####################

```{r}
tbl(con, 'planes_distance') %>%
  inner_join(tbl(con, 'planes'), by='tailnum') %>%
  arrange(desc(total_distance)) %>%
  select(total_distance, manufacturer, model)
```  
  

10. Check what is the actual SQL query generated by the code you have created in exercise 9.
 

####################
#                  #
#    Exercise 10   #
#                  #
####################

```{r}
tbl(con, 'planes_distance') %>%
  inner_join(tbl(con, 'planes'), by='tailnum') %>%
  arrange(desc(total_distance)) %>%
  select(total_distance, manufacturer, model) %>%
  show_query()
```








